Analysing the 2024 U.S. Presidential Election

### 2024-11-04 paper/paper.qmd
Edit README.md and change of the paper's formating
Update paper: Main body section

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - David Qi
  - Haobo Ren
  - Xinrui Xie
thanks: "Code and data are available at: https://github.com/HaoboRrrr/USA_Election"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
```

# Introduction

# Data {#sec-data}

## Overview
Josh Pasek have discussed how to the use of Aggregation and Predictive Modeling on polling can increase prediction effectiveness[@Pasek_2015]. In this article, we will make use of predictive modeling to create a model that aim to predict the support rate of Kamala Harris in the 2024 United States presidential election using a linear regression model trained form past polling results and Dow Jones Industrial Average.

## Measurement

## Outcome variables

## Predictor variables

# Model

We will use a linear model to predict the general polling result of Harris.

## Model set-up
For each poll, we assumed they are being sampled similarly. \
Let the support rate for Harris at day $t$ be denoted by $R_t$. And we denote the Dow Jones Industrial Average at day $t$ by $D_t$.
Then we will use linear regression model: 
$$
R_t = D_{t-60} +  D_{t-90} + \epsilon_t
$$
Where $\epsilon_t$ is an independent, normally distributed error term.

## Model justification

We expect a positive relationship between the support rate of Harris and Dow Jones Industrial Average 60 and 90 days before. Since people tend to support the party in power when the economy is better, where the condiction of the economy is reflected by $D_t$.

TODO: find paper to support this.
finishing reading https://dspacemainprd01.lib.uwaterloo.ca/server/api/core/bitstreams/4dfe49d7-72b7-44e5-8fdc-8a5cf0b77b24/content

# Results

# Discussion

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

@@

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

## Diagnostics

@@

# Appendix 1: Polling methodology

@@

>>>>

---
title: "Analysing the 2024 U.S. Presidential Election"
subtitle: "Spotlight on North Carolina's Role as a Battleground State"
author: 
  - David Qi
  - Haobo Ren
  - Xinrui Xie
thanks: "Code and data are available at: https://github.com/HaoboRrrr/USA_Election"
date: today
date-format: long
abstract: "This paper examines the relationship between polling results and economic conditions to forecast the 2024 U.S. presidential election, with results analyzed for both swing states and at a national level. Findings suggest that Republicans have a modest chance (6.6%) of winning, with North Carolina emerging as the pivotal swing state. The study may offer guidance for campaign strategies and market forcasts. While providing insight on changes of swing states in U.S. politics. "
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

library(tidyverse)
library(arrow)
library(knitr)
library(ggplot2)

harris_data <- read_parquet(here::here("data/02-analysis_data/just_harris_high_quality.parquet")) %>% select(pct, start_date,end_date,question_id,state)
trump_data <- read_parquet(here::here("data/02-analysis_data/just_trump_high_quality.parquet")) %>% select(pct, start_date,end_date,question_id,state)
DJIA_data = read_parquet(here::here("data/02-analysis_data/DJIA_filled.parquet"))

merged_summary = read_parquet(here::here("data/02-analysis_data/merged_summary.parquet"))

model_harris = readRDS(file = here::here("models/model_harris.rds"))

model_trump = readRDS(file = here::here("models/model_trump.rds"))

prediction_results = read_parquet(here::here("models/prediction_results.parquet"))

trump_data <- read_parquet(here::here("data/02-analysis_data/just_trump_high_quality.parquet"))

```

# Introduction {#sec-intro}

Josh Pasek have discussed how to the use of Aggregation and Predictive Modeling on polling can increase prediction effectiveness[@Pasek_2015]. Following this principle, we will make use of predictive modeling to create a model that aim to predict the results of the 2024 United States presidential election. This paper uses polling data gathered by FiveThirtyEight[@fivethirtyeight] with Dow Jones industrial average data[@sp_dow_jones_2024] as an indicator of the economy to train a simple linear regression model to predict the result of election in each swing state.

The model aims to predict the result of a poll that ends on November $5^{th}$, 2024, we believe this will reflect the result of the upcoming election on that day, the result of the election may be assumed to be close to the mean of these polls. And by this process we derive the result of the election. After the result is gained, we computed the electoral votes for the state, and used Monte Carlo simulation to derive winning possibility of each party.

We conclude that Donald Trump have $6.6\%$ probability of winning this election. From our prediction, we highlight North Carolina as the most important battleground state. The state has a predicted $14.2\% $probability to be won by Trump, and will give him 16 electoral votes. Along with the votes that we predict he will almost certainly gain, this pushes him one step (9 electoral votes) to victory. This provide power insight to how campaign and fund raising strategies should be targeted on the election day. And provides a forecast on one of the most important influences in economy in the next several months, which will be especially helpful in planning market stragities.

The scripts in this paper are written in R[@citeR], The table in our study are generated by knitr [@knitr],  and graph with ggplot2[@ggplot2]. Data input and output were done using arrow[@arrow], and curl[@curl]. Data cleaning made use of janitor[@janitor].

The rest of the paper is structured as follows: in [Data](#sec-data) section we will discuss the various aspects of the data we use, how they were gathered and the variable we shall use. In the [Model](#sec-model) section we will discuss details of our model, why it is justified and the coefficient summaries.  In the [Results](#sec-results) section, we present the predictions of the model. In the [Discussion](#sec-discussion) section we will discuss the interpretation to the predictions, and what we may learn from the coefficients for the model, we will also look as possible weakness and future improvement for the model.

# Data {#sec-data}
We will make use of polling data organized by FiveThirtyEight[@fivethirtyeight], collected from polling made by various sources before the election. As well as the Dow Jones industrial average data[@sp_dow_jones_2024].

## Overview For Polling data
Fist we take a look at the the polling data.

The data set contain 17208 observations of polls, for our propose, we choose only to include 1176 entries, the justification for this data selection is given in the [Justification](#justification_poll) section.

##Data Measurement
The data was organized by FiveThirtyEight (2024), collected form polls made by various institutions, they each use a distinct methodology. While FiveThirtyEight checks their methodology, and include the polling data if it meets their standard. An example of the methodology of one poll is included in [Appendix 1](#appendix1) It is a reliable and comprehensive source of US election polls， so we have choose this data set.

## Justification For Poll Selection{#justification_poll}
The goal of our study is to predict the result of US presidential election, it is known that the only reasonable winner will be form the Democratic Party, or the Republican Party. Which can be seen form their overwhelmingly high support rate in the polling results and historical results. So we will only include the polling results for Kamala Harris and Donald Trump, the candidate for the two parties.

Another observation coming form past experience that is not so obvious is that the result of presidential elections are often only determined by the result in "swing states".[@why_only_ten; @aldrich2023change] Following the definition given by Schultz and Hecht, which have given having below 5% difference between the two parties as one criteria, we have examined our data and have identified the swing states from the data set, past experience, and outside sources.

[@tbl-polloverview] Shows the absolute value of the difference in average support rate from all the polls. If we take all states such the absolute difference is less than 5% to be swing states, similar to the method of less than 5% difference between two parties in the last election. We will see they coincide except that our list includes Maine CD-2 and have not included Florida[@fec2020federalelections]. This list, when Florida is added, also include the swing state listed by popular media[@bbc2024election;@dorn2024election;@abcnews2024trumpleads] and polling platform[@fivethirtyeight2024swing] before this upcoming election.

Also note that some states only have very limited polls, which will make modeling these states very difficult and inaccurate, the swing states we choose allow us to have sufficient polls for each state.

The list of swing states we use are: North Carolina, Pennsylvania, Nevada, Georgia, Arizona, Michigan, Wisconsin, Maine's 2nd congressional district, and Florida. We will also include national polls to help with modeling.

The data have also contained ratings for each poll based on their methodology and transparency. As a summary made to understand the data, [@tbl-polloverview] is made by selecting all the polling, however, for the rest of this study, we will only use polls that are rated 2.5 or above by FiveThirtyEight to ensure quality of data.

```{r}
#| label: tbl-polloverview
#| tbl-cap: Comparison of average polling result for two candiates
#| include: true
#| echo: false
#| warning: false
#| message: false

merged_summary = merged_summary %>% select(state,total_count.x,mean_pct_harris,mean_pct_trump,abs_diff)  %>% arrange(abs_diff)

kable(merged_summary, col.names = c("State", "Number of Polls" , "Average Support Rate(Harris)","Average Support Rate(Trump)","Absolute difference"))
```

## Varibles
This section explains the important variables relevant to our analysis: Percent, End Date, state, numeric grade, Candidate Name and sample size.

### Percent
Percent (`pct`) is the percentage support rate learned by this poll. This is the main estimated we are targeted to model, the justification of using this as estimand is given in model section.

### End Date
End Date(`end_date`) is a time variable that specify the date that the poll ends.

### State
State (`state`) is a categorical variable representing each U.S. state or Congressional district that is considered a separate entity in the U.S. presidential election. Each poll reflect the result of a state, if the state is not specified, the poll is national.

### Numeric grade
Numeric grade (`numeric_grade`) variable is a measurement of the quality of the poll given by FiveThirtyEight.It is a numeric score from 3 to 0. In this study, we will only use polls with Numeric grade higher than 2.5.

### Candidate Name 
Candidate Name (`candidate_name`) is a categorical variable that specify the name of candidate whose the support rate this poll measure.

### Sample Size
[@fig-samplesize] is a plot showing the sample size(`sample_size`) of the data we use for modeling. The sample size of each poll is 432 minimum and 78247 maximum, the mean sample size is 1497. Observe that the sample size is extremely small comparing to the entire population of a state, this observation is important in how we predict using our model.

```{r}
#| label: fig-samplesize
#| tbl-cap: Comparison of average polling result for two candiates
#| include: true
#| echo: false
#| warning: false
#| message: false

trump_data = trump_data %>% filter(sample_size < 10000)

ggplot(trump_data, aes(x = sample_size)) +
  geom_histogram( binwidth = 200, fill = "steelblue", color = "white", alpha = 0.7,xlim = 10000)+
  labs(x = "Sample size",
       y = "Frequency") +
  theme_minimal()
```

## Overview Dow Jones Data
The Dow Jones Industrial Average(DJIA) is a renowned economic indicator that launched in 1896[@spglobal_djia]. It is used long in economical analysis and is considered "one of the analyst's oldest friend and one of his most useful tool"[@milne1966dow]. The index is still being used in economic studies today and is considered an important indicator of the economic of the United States[@arendas2018halloween].

The data includes average DJIA data for each weekday, there are no data for weekends and certain holidays that DJIA do not operate. For the propose of this analysis, we have filled the data by assigning these dates with DJIA values on the closest date before. The data after filling is a numerical value corresponding to each day.

## DJIA Data Measurement
The data publisher(S&P Dow Jones Indices LLC) have full transparency and reproducible of their methodology[@milne1966dow], DJIA is a weighted average of stock price of 30 firm that are considered important to all aspect of economy except transportation and utilities. The firm and corresponding weights are fully transparent. As these firms play a key role in the U.S. economy and the presidential election, Dow Jones Industrial Average will be a useful predictor for electoral results.

It is also a daily updated, easy to access data that is not subject to change. Some other important indicators of economy, such as United States Nonfarm Payrolls, only updates monthly and previous data may be corrected in the future(@bls_employment_situation).

# Model {#sec-model}

We will use simple linear regression model to predict the general polling result of Harris and Trump. The mode will be implemented using R[@citeR].

## Model set-up
For each poll, we assumed they are being sampled similarly. \
Let the support rate for Harris in a single poll that end at day $t$ be denoted by $H_t$. And we denote the Dow Jones Industrial Average at day $t$ by $D_t$.Denote State by $s$, and the corresponding regression coefficient by $\beta_s$
Then we will use linear regression model: 
$$
H_t = \beta_0 + \beta_1D_{t-7} +  \beta_2D_{t-14}+\beta_3D_{t-28}+\beta_4D_{t-60}+\beta_5D_{t-150} +\beta_6D_{t-180}+ \beta_s + \epsilon_t
$$
Where $\epsilon_t$ is an independent, normally distributed error term.  \
$\beta_0,\beta_1,\beta_2,\beta_3,\beta_4,\beta_5,\beta_6$ are regression coefficients. \
$\beta_s$ is the corresponding regression coefficient to state $s$. \
We will predict the election support rate of Harris to be $mean(H_t)$ \

For Donald Trump, we use a similar model, let the support rate in a single poll for Trump at day $t$ be denoted by $T_t$. The the linear regression model can be written as: 
$$
T_t = \alpha_0 + \alpha_1D_{t-7} +  \alpha_2D_{t-14}+\alpha_3D_{t-28}+\alpha_4D_{t-60}+\alpha_5D_{t-150} +\alpha_6D_{t-180}+ \alpha_s + e_t
$$
Where $e_t$ is an independent, normally distributed error term.  \
$\alpha_0,\alpha_1,\alpha_2,\alpha_3,\alpha_4,\alpha_5,\alpha_6$ are regression coefficients. \
$\alpha_s$ is the corresponding regression coefficient to state $s$. \
We will predict the election support rate of Donald Trump to be $mean(T_t)$ \

## Model justification
In deciding this model, we have made many decisions. In this section we justify the importat choices we made during modeling.

### Use of DJIA
The use of economic data to predict result in presidential election can be supported by various studies that shows the result and voting pattern in presidential elections are highly correlated to the economy[@aldrich2023change]. And DJIA is an important and wide used indicator of the economy. We hypothesize that DJIA will have a linear relationship with the support rate. We expect the effect economy has on polling results to have a delayed effect, which is the reason for including lagged terms in linear regression.

An alternative approach is to use changes in the DJIA to reflect economic fluctuations. However, we choose to use the absolute DJIA values, as inputing missing days for change data is difficult to justify, and the absolute economic status provides essential insights. Our hypothesis is that a strong economy favors the ruling party in elections, as switching parties could risk economic shifts and potentially lower living standards. Here, the absolute economic level serves as an indicator, reflecting the overall standard of living.

### Use of mean to predict election {#usemean}
We may treat each poll as an observation for the population of support rate, what our model does is that it predicts the Percent('pct') variable in a single poll. Since the sample size of is extremely small comparing to the population of each state, which is 1497 people on average comparing to at least 700 thousand people. And by statistic around 60% people will participate in each election[@aldrich2023change], we may consider the election result to be very close to the population mean by the weak law of large numbers. So by considering the mean of the predicted polls, we may predict the election result.

### Using different model for Harris and Trump
There exist the possibility for using one model with a categorical variable, however we find the voting behavior of the two groups of people may be vastly different and cannot be summarized by a simple shift in the regression line. The use of interaction terms may capture this difference, however will make the model more complicated and difficult to write out.

### Using the same model for each state
As states in United States, we expect each state to share some similarity with other states, as well as with the national data. So we have made state a categorical variable in regression. The main difference with candidates here is that we have data on national polls, which have direct correlation with the polling result of each state rather than just hypothesized connection, to make use of those data, we include them by using a categorical for state.

### Linerity and Independence Assumption
The justification for these assumptions are in [Appendix 3](#appendix3)

### Reason for choosing these lag intervals
Selection process is explained in [Appendix 3](#appendix3), we have explicitly made two model useing the same set of variables so comparison is possible.

### Including only swing states
As we stated above, the election is determined almost solely by swing states. And many other states do not have enough data for sound prediction. Thus our model only include the data for swing states and nation polls.

## Model summary
[@tbl-modelharris] and [@tbl-modeltrump] are the model coefficients fitted for Harris and Trump. The model for Harris have adjusted R-squared 0.1781, and model for trump have adjusted R-squared 0.3195.

Note that DJIA_lag_s is the variable $D_{t-s}$

```{r}
#| label: tbl-modelharris
#| tbl-cap: Model summary for Harris
#| include: true
#| echo: false
#| warning: false
#| message: false

model_summary_harris <- summary(model_harris)

coef_table_harris <- as.data.frame(model_summary_harris$coefficients)

kable(coef_table_harris)
```

```{r}
#| label: tbl-modeltrump
#| tbl-cap: Model summary for Trump
#| include: true
#| echo: false
#| warning: false
#| message: false

model_summary_trump <- summary(model_trump)

coef_table_trump <- as.data.frame(model_summary_trump$coefficients)

kable(coef_table_trump)
```

# Results {#sec-results}

## Model prediction

[@tbl-predictionharris] shows the predicted result for the election in the swing states as well as the national result. However, this table have not considered confidence intervals.
Note that the reason for using confidence interval instead of prediction intervals is that as explained in [Use of mean to predict election](@usemean) section, the actual result of the election is expected to be close to the population mean. 
For instance in North Carolina, the 95% confidence interval for Trump is $[48.22990, 49.91708]$, while for Harris it is $[48.90044, 50.54273]$, they have overlapped, which means that we are not very certain with the result. To gain a better understanding of the possibility of a person winning in each state and the nation, we use Monte Carlo simulation to get the possibility. The last column of [@tbl-predictionharris] is made using this method and give the possibility we predict for Trump to win each state.

```{r}
#| label: tbl-predictionharris
#| tbl-cap: Prediction for Each state
#| include: true
#| echo: false
#| warning: false
#| message: false

graphing_result = prediction_results %>% select(state, prediction_trump,prediction_harris,result,trump_win_rate)

kable(graphing_result, col.names = c("State" , "Prediction Trump" ,"Prediction Harris", "Predicted Result","Trump's Win Rate"))
```

Note that the United States uses Electoral College system to determine the winner of the election. Given this model, we may predict using Monte Carlo simulation the win rate of each candidate. Assume that each state except for swing states voted for the the party they voted for in 2020, this gives Democratic party 226 electoral votes and The republican 188 electoral votes. Using our model, we find that republican wins 66199 out of 1000000 simulation with no draws, mean electoral votes for democratic is 284.765953 and mean electoral votes for republican is 253.234047. The estimated win rate for republican is 0.066199. If we follow our prediction with out considering possibility, Harris will win by 293 versus 245 electoral votes.

# Discussion{#sec-discussion}
Our studies have presented a forecast of the U.S. 2024 presidential election

## Interpretation of Model Coefficients
First we note that unlike our initial instinct, the votes of the two parties are not always mutually exclusive. In the case of $\beta_6$ and $\alpha_6$(DJIA_lag_180), the two candidates both have positively correlated support rate. This may indicate that people are less likely to select an independent candidate when the economic is good, as people will be less likely to seek change under such circumstance. However in the case of states, we see that every $\beta_s$ have opposite sign from $\alpha_s$.

We also see there is a difference in the R-squares, this may tell us that economy is a better predictor for support rate of Trump than Harris, possibly indicating that Trump has a campaign more focused on economy. $\beta_6$ is almost 2 times as large as $\alpha_6$, which aligns with our hypothesis that good long term economy will make the election favor the ruling party.

## Comparson with 2020 
Looking back at 2020, based on our prediction, the difference was that Harris is predicted to take North Carolina and the entire Maine, While Trump is predicted to win Arizona and Georgia, which makes the republican win 10 more votes than the 2020 election. Or 13 more votes after considering the slight changes to number of electoral votes of each state. However the Democratic party is in a great advantage and will not be defeated unless Trump wins in North Carolina. Which by our prediction is difficult but not impossible.

## Impact of Each Swing state
The modeling have highlighted the impact swing states has on presidential election. And also suggest the definition of swing state is subject to change. Florida may be considered a swing state from the election results in 2020, however the polling result and model analysis have shown that Trump will certainly win this state. We also need to highlight the importance and closeness of the election in North Carolina. As a swing state with 16 electoral votes, we predict Republican Party to have a reasonable probability to win North Carolina, and North Carolina will be deceive in determining the election results.

## Weaknesses and next steps
Our studies have relied solely on DJIA data and have not considered other factors that also has great influence to the result of the presidential election. Although as Pasek(2015) says, it is never possible to model all the factors, it is even unrealistic to model all the important factors since each year's election is unique in it's way. We still may improve the analysis by giving consideration to more variables. For instance, we may do a weighted regression by using the numerical scores as weights, or consider other measurements of economy that covers the transportation and utility aspect of economy that DJIA did not cover.

Although we have claimed that non-swing states will almost certainly vote for the party they supported, bringing in their data still may help our. However we have to note having all the states will greatly increase the complexity of the model, and prediction for many of the states will not be trustworthy due to scarcity of relevant data.

In the modeling, we have view polling as taking one sample form the population, while election as taking 60% of the data form the population. However, polls may be biased by various factors.For example non-response bias or moderacy bias[@stantcheva2023how]. A closer and more careful investigation by looking at the difference between poll results an actual election outcomes is required to understand the relationship between polls and actual election.

@@

@@

# Appendix 1: Polling methodology {#Appendix1}

@@

# Appendix 3: Model diagnostics {#appendix3}

[@fig-diagnostics_harris] and [@fig-diagnostics_trump] are the model diagnostics for the two models. We can see that the residue only have limited deviation from normality and did not display an obvious pattern, which justifies the assumption of linearity and independence.

```{r}
#| label: fig-diagnostics_harris
#| tbl-cap: Model diagnostics for Harris model
#| include: true
#| echo: false
#| warning: false
#| message: false
par(mfrow = c(2, 2))  
plot(model_harris)
```

```{r}
#| label: fig-diagnostics_trump
#| tbl-cap: Model diagnostics for Trump model
#| include: true
#| echo: false
#| warning: false
#| message: false
par(mfrow = c(2, 2))  
plot(model_trump)
```
## Selection

Our model selection is based on AIC(Akaike information criterion) and interpretation of data.

The best model in terms of AIC in selecting was "DJIA_lag_7 + DJIA_lag_14 + DJIA_lag_90 + DJIA_lag_120 + DJIA_lag_150 + DJIA_lag_180 + state" for Harris and "DJIA_lag_14 + DJIA_lag_28 + DJIA_lag_60 + DJIA_lag_90 + DJIA_lag_150 + DJIA_lag_180 + state" For Trump. However, we want the two model to have the same set of predictors, which may increase interpretability of the model. So we have selected a model that include relevant terms for both candidates.
